---
title: 'PS5: rmarkdown and plotting'
author: "Alexia Chan"
date: "2023-02-15"
output: html_document
---

# 1.1
```{r, message=FALSE}
library(knitr)
library(tidyverse)
data <- read.delim("gapminder.csv.bz2")
```

# 1.2

##### There are `r nrow(data)` rows and `r ncol(data)` columns

# 1.3
```{r}
sample_n(data, 3) %>% 
  knitr::kable()
```

# 2.1
##### Displays the number of different countries that each variable contains
```{r, message=FALSE}
data %>% 
  select(name, iso3, iso2) %>% 
  summarise(name = n_distinct(name), iso3 = n_distinct(iso3), iso2 = n_distinct(iso2))
```

# 2.2 (a)
##### Iso-2 codes that correspond to more than 1 name
```{r, message=FALSE}
data %>% 
  select(iso2, name) %>% 
  group_by(iso2) %>% 
  summarise(unique = n_distinct(name)) %>%   
  filter(unique > 1) 
```

yes, there is one iso-2 code that has two names that correspond to it

# 2.2 (b) 
##### Countries that correspond to more that 1 iso-3 code
```{r message=FALSE}
data %>% 
  group_by(name) %>% 
  select(iso3, name) %>% 
  summarise(unique = n_distinct(iso3)) %>% 
  filter(unique > 1)
```

yes, there is one name that has four iso-3 codes that correspond to it

# 2.3
##### Minimum and maximum year in the data
```{r}
data %>% 
  select(time) %>% 
  filter(!is.na(time)) %>% 
  summarize(min = min(time), max = max(time))
```

# 3.1
##### Missing CO2 and CO2 PC emissions per year
displays the top few countries with the largest number of missing CO2 data
```{r}
data %>%
  group_by(time) %>% 
  select(time, co2, co2_PC) %>% 
  summarise(co2 = sum(is.na(co2)), co2_PC = sum(is.na(co2_PC)), sum = co2 + co2_PC) %>% 
  arrange(-sum) %>% 
  head()
```
years 2017-2019 have the most missing data

# 3.2
##### Plot shows total CO2 emissions over time for each region
```{r message=FALSE, warning=FALSE}
data %>% 
  filter(iso3 == c("USA", "CHN", "IND", "JPN", "ITA")) %>% 
  ggplot() +
    geom_line(aes(x = time, y = co2, col = iso3)) +
    geom_line(aes(x = time, y = co2_PC, col = iso3)) +
    ggtitle("Total CO2 Emissions Over Time") +
    xlab("Years") +
    ylab("Total CO2")
```

- Most countries increase their CO2 emissions over time
- China has the highest, fastest increase out of all of them
- The USA is the only country where CO2 starts to go down

# 3.3
##### Plot of CO2 emissions per capita over time for each country
```{r message=FALSE, warning=FALSE}
data %>% 
  filter(iso3 == c("USA", "CHN", "IND", "JPN", "ITA")) %>% 
  ggplot() +
    geom_line(aes(x = time, y = co2_PC, col = iso3)) +
    ggtitle("CO2 Emissions per Capita") +
    xlab("Years") +
    ylab("CO2 per Capita")
```

- This suggests that the USA produces the most CO2, and India produces the least

# 3.4 
##### Average CO2 emissions per capita for each continent in 2016

```{r message=FALSE}
data %>% 
  group_by(region) %>%
  filter(time == 2016) %>% 
  select(region, co2_PC) %>% 
  filter(!is.na(co2_PC), nzchar(region)) %>% 
  summarise(avg = mean(co2_PC))
```

- Europe produces the most CO2 on average
- Africa produces the least CO2 on average

# 3.5
##### Average CO2 emissions per capita for each continent in 1960 and 2016
```{r message=FALSE, warning=FALSE}
years <- c(1960, 2016)

databarplot <- data %>%
  filter(time %in% years) %>% 
  group_by(region, time) %>%
  #select(region, co2_PC) %>% 
  filter(!is.na(co2_PC), nzchar(region)) %>%
  summarise(avg = mean(co2_PC), groups = "keep") 

ggplot(data = databarplot) +
  geom_col(aes(x = region, y = avg, fill = as.factor(time)), stat = "identity", pos = position_dodge2(0)) +
  ggtitle("Average CO2 Emissions per Capita") +
  xlab("Region") +
  ylab("Average CO2 Emissions per Capita") +
  labs(fill = "Year")
```

# 3.6
##### Top 3 largest and smallest countries in terms of CO2 PC emission in 2016
(2019 has many NA values)
```{r message=FALSE}
dataset <- data %>% 
  filter(time == 2016, !is.na(co2_PC), nzchar(name)) %>% 
  select(name, co2_PC) %>%
  group_by(name) %>%
  arrange(-co2_PC)

# Top 3 countries with the largest CO2 PC emission
dataset %>% 
  head(3)
# Top 3 countries with the smallest CO2 PC emission
dataset %>% 
  tail(3)
```

# 4.1
##### Plot of GDP per capita and life expectancy per country in 1960
```{r message=FALSE, warning=FALSE}
data %>% 
  filter(time == 1960, nzchar(region)) %>% 
  ggplot() +
    geom_point(aes(x = GDP_PC, y = lifeExpectancy, col = region, size = totalPopulation)) +
    ggtitle("GDP per Capita vs Life Expectancy by Country in 1960")
    
```

- Countries with high GDPs tend to have high life expectancies
- The largest regions have the lowest GDP, but in general, size doesn't seem to have a large effect 
- Europe has the highest life expectancy and GDP
- Africa has the lowest life expectancy and GDP

# 4.2
##### Plot of GDP per capita and life expectancy per country in 2019
```{r message=FALSE, warning=FALSE}
data %>% 
  filter(time == 2019, nzchar(region)) %>% 
  ggplot() +
    geom_point(aes(x = GDP_PC, y = lifeExpectancy, col = region, size = totalPopulation)) +
    ggtitle("GDP per Capita vs Life Expectancy by Country in 2019")
```

- Africa has the lowest GDP and life expectancy
- China has the largest total population but a low GDP

# 4.3
##### Comparison between plots above
- Each country's ranking is relatively the same since 1960
- Life expectancy overall has increased by about 10 years
- GDP overall has increased 

# 4.4
##### Average life expectancy per continent in 1960 and 2019
```{r message=FALSE, warning=FALSE}
data %>%
  select(region, lifeExpectancy, time) %>%
  group_by(time, region) %>% 
  filter(!is.na(lifeExpectancy), nzchar(region)) %>% 
  filter(time == c(1960, 2019)) %>% 
  summarise(average = mean(lifeExpectancy))
```

The results did fit with what I see on the figures

# 4.5 
##### Average life expectancy growth from 1960-2019 per region
```{r message=FALSE, warning=FALSE}
data %>% 
  filter(time == c(1960, 2019), !is.na(lifeExpectancy), nzchar(region)) %>% 
  select(region, lifeExpectancy, time) %>% 
  group_by(region, time) %>% 
  summarise(average = mean(lifeExpectancy)) %>% 
  mutate(previous = lag(average), growth = average - previous) %>% 
  select(region, growth) %>% 
  filter(!is.na(growth))
```

# 4.6
##### Plot of GDP PC for 1960 and 2019
```{r message=FALSE}
years <-  c(1960, 2019)
graph <-  data %>% 
  filter(time %in% years, !is.na(GDP_PC)) %>% 
  group_by(time) %>% 
  select(GDP_PC, time) 

ggplot(data = graph) +
  geom_histogram(aes(x = GDP_PC, fill = as.factor(time)), position = "dodge")+
  ggtitle("GDP Per Capita for 1960 and 2019")
```

# 4.7 
##### Ranking of US in terms of life expectancy in 1960 and 2019
```{r message=FALSE, warning=FALSE}
data %>% 
  select(name, lifeExpectancy, time) %>% 
  filter(time == 1960, !is.na(lifeExpectancy), nzchar(name)) %>% 
  arrange(-lifeExpectancy) %>% 
  mutate(rank = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America")
data %>% 
  select(name, lifeExpectancy, time) %>% 
  filter(time == 2019, !is.na(lifeExpectancy), nzchar(name)) %>% 
  arrange(-lifeExpectancy) %>% 
  mutate(rank = rank(-lifeExpectancy)) %>% 
  filter(name == "United States of America")
```

The US is ranked 17 in terms of life expectancy in 1960, and 46 in 2019

# 4.8 
##### Relative rank divided by the corresponding number of countries in that year
```{r message=FALSE, warning=FALSE}
countries1960 <- data %>% 
  filter(time == 1960, !is.na(lifeExpectancy), nzchar(name)) %>% 
  summarise(countries = n())
countries2019 <- data %>% 
  filter(time == 2019, !is.na(lifeExpectancy), nzchar(name)) %>% 
  summarise(countries = n())

data %>% 
  filter(time == 1960, !is.na(lifeExpectancy), nzchar(name)) %>%
  select(name, lifeExpectancy) %>% 
  arrange(-lifeExpectancy) %>% 
  mutate(rank = rank(-lifeExpectancy), new = rank / 188) %>% 
  filter(name == "United States of America")
data %>% 
  filter(time == 2019, !is.na(lifeExpectancy), nzchar(name)) %>%
  select(name, lifeExpectancy) %>% 
  arrange(-lifeExpectancy) %>% 
  mutate(rank = rank(-lifeExpectancy), new = rank / 196) %>% 
  filter(name == "United States of America")
```

I probably spent about 8.5 hours on this assignment


